---
title: "Preprocessing audit summary"
subtitle: "04__DataPreProcess__Summary — how uncleaned becomes cleaned"
author: "Michael Sieler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
# Filename: 04__DataPreProcess__Summary.Rmd
# Created by: Michael Sieler
# Date last updated: 2026-04-22
#
# Description:
#   Audit report for Sieler2026 preprocessing. Compares the latest uncleaned phyloseq
#   (from Data/DADA2) to the latest cleaned phyloseq produced by 04__DataPreProcess.R.
#   Replays the major filter steps (including optional decontam) to show which samples
#   and taxa are removed at each stage.
#
# Expected input:
#   - Sieler2026 repo with Data/DADA2/pseq_uncleaned_*.rds and pseq_cleaned_filtered_*.rds
#   - Optional: Data/r_objects/decontam__*.rds (written by 04__DataPreProcess.R when run_decontam=TRUE)
#
# Expected output:
#   - HTML audit report (this document)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

# Knit from any working directory: locate repo root, then source_project_init.
start <- if (requireNamespace("knitr", quietly = TRUE) && !is.null(knitr::current_input())) {
  dirname(knitr::current_input(dir = TRUE))
} else {
  getwd()
}
init_rel <- file.path("Code", "00__Setup", "source_project_init.R")
root <- normalizePath(start, mustWork = TRUE)
while (!file.exists(file.path(root, init_rel))) {
  parent <- dirname(root)
  if (identical(parent, root)) {
    stop("Could not find ", init_rel, " above ", start)
  }
  root <- parent
}
source(file.path(root, init_rel))
sieler2026_source_initialize_project(root = root)
knitr::opts_knit$set(root.dir = root)
```

## Load the latest phyloseq objects (repo-local)

```{r}
ps_uncleaned <- readRDS(path.pseq.uncleaned)
ps_cleaned <- readRDS(path.pseq.cleaned)

tibble::tibble(
  object = c("uncleaned (latest)", "cleaned (latest)"),
  path = c(path.pseq.uncleaned, path.pseq.cleaned),
  n_samples = c(phyloseq::nsamples(ps_uncleaned), phyloseq::nsamples(ps_cleaned)),
  n_taxa = c(phyloseq::ntaxa(ps_uncleaned), phyloseq::ntaxa(ps_cleaned))
) |>
  gt::gt() |>
  gt::tab_header(title = "Current preprocessing inputs/outputs (latest RDS)")
```

## Read-depth + taxa filter settings used by preprocessing

These defaults are defined in `Code/00__Setup/04__DataPreProcess.R`.

```{r}
min_prevalence_taxa <- 3L
min_sample_reads <- 5000L
```

## Optional decontam audit (if present)

If you ran `04__DataPreProcess.R` with decontam enabled, it writes a dated audit RDS under `Data/r_objects/`.

```{r}
decontam_audit_path <- resolve_latest_rds(
  dir = path.objects,
  pattern = "^decontam__.*\\.rds$",
  error_if_empty = FALSE
)

if (!is.na(decontam_audit_path)) {
  decontam_audit <- readRDS(decontam_audit_path)
  gt::gt(tibble::tibble(
    key = names(decontam_audit)[names(decontam_audit) != "contam_tbl"],
    value = purrr::map_chr(decontam_audit[names(decontam_audit) != "contam_tbl"], \(x) paste0(x, collapse = ", "))
  )) |>
    gt::tab_header(
      title = "Decontam audit file detected",
      subtitle = basename(decontam_audit_path)
    )

  if (!is.null(decontam_audit$contam_tbl) && nrow(decontam_audit$contam_tbl) > 0) {
    decontam_audit$contam_tbl |>
      dplyr::filter(.data$contaminant) |>
      dplyr::mutate(p = round(.data$p, 4)) |>
      dplyr::select(.data$taxon, .data$contaminant, .data$p) |>
      gt::gt() |>
      gt::tab_header(
        title = "Taxa flagged as contaminants by decontam",
        subtitle = "decontam::isContaminant(method = 'prevalence')"
      )
  }
} else {
  gt::gt(tibble::tibble(note = "No decontam audit file found in Data/r_objects/. Decontam may not have been run yet.")) |>
    gt::tab_header(title = "Decontam audit")
}
```

## Recreate the filtering steps (to audit what was removed)

This section replays the same logic as `apply_pre_analysis_filters()` to show **which** taxa/samples are removed by each stage.

```{r}
recreate_pre_analysis_filters <- function(ps_input, min_prevalence, min_reads) {
  # Step 0: start
  ps0 <- ps_input

  # Step 1: prevalence (taxa)
  otu <- as(phyloseq::otu_table(ps0), "matrix")
  if (!phyloseq::taxa_are_rows(ps0)) {
    otu <- t(otu)
  }
  taxa_prev <- rowSums(otu > 0)
  ps1 <- phyloseq::prune_taxa(taxa_prev >= min_prevalence, ps0)

  # Step 2: taxonomy-based taxa removal (if available)
  ps2 <- ps1
  if (!is.null(phyloseq::tax_table(ps2, errorIfNULL = FALSE))) {
    tax_df <- as.data.frame(as(phyloseq::tax_table(ps2), "matrix"), stringsAsFactors = FALSE)
    rm_target <- apply(
      tax_df,
      1,
      function(x) any(grepl("Mitochondria|Chloroplast|Eukaryota", x, ignore.case = TRUE))
    )
    phylum_col <- if ("Phylum" %in% colnames(tax_df)) {
      "Phylum"
    } else if (ncol(tax_df) >= 2) {
      colnames(tax_df)[2]
    } else {
      NA_character_
    }
    unresolved <- rep(FALSE, nrow(tax_df))
    if (!is.na(phylum_col)) {
      phylum_vals <- tax_df[[phylum_col]]
      unresolved <- is.na(phylum_vals) |
        phylum_vals == "" |
        grepl("^Kingdom_Bacteria$", phylum_vals, ignore.case = TRUE) |
        grepl("^Bacteria Phylum$", phylum_vals, ignore.case = TRUE)
    }
    keep <- !(rm_target | unresolved)
    ps2 <- phyloseq::prune_taxa(rownames(tax_df)[keep], ps2)
  }

  # Step 3: low-read samples (after taxa filters)
  keep_samples <- phyloseq::sample_sums(ps2) >= min_reads
  ps3 <- phyloseq::prune_samples(keep_samples, ps2)
  ps3 <- phyloseq::prune_taxa(phyloseq::taxa_sums(ps3) > 0, ps3)

  list(
    ps0 = ps0,
    ps1 = ps1,
    ps2 = ps2,
    ps3 = ps3,
    keep_samples_step3 = keep_samples
  )
}

steps <- recreate_pre_analysis_filters(
  ps_input = ps_uncleaned,
  min_prevalence = min_prevalence_taxa,
  min_reads = min_sample_reads
)

tibble::tibble(
  step = c("0 raw (uncleaned)", "1 prevalence taxa", "2 taxonomy taxa", "3 low-read samples"),
  n_samples = c(phyloseq::nsamples(steps$ps0), phyloseq::nsamples(steps$ps1), phyloseq::nsamples(steps$ps2), phyloseq::nsamples(steps$ps3)),
  n_taxa = c(phyloseq::ntaxa(steps$ps0), phyloseq::ntaxa(steps$ps1), phyloseq::ntaxa(steps$ps2), phyloseq::ntaxa(steps$ps3))
) |>
  gt::gt() |>
  gt::tab_header(title = "Dimensions after each preprocessing filter stage")
```

### Samples removed for low read depth

```{r}
reads_after_tax_filters <- phyloseq::sample_sums(steps$ps2)
low_read_removed <- names(reads_after_tax_filters)[reads_after_tax_filters < min_sample_reads]

sd <- phyloseq::sample_data(ps_uncleaned) |>
  as.data.frame() |>
  tibble::rownames_to_column("sample_name") |>
  tibble::as_tibble()

# Robust metadata lookup: some datasets store the fecal ID in a column rather than the rowname,
# or mix "153" vs "f153" conventions. Build a mapping table from any available identifier column.
id_cols <- intersect(
  c("sample_name", "fecal.sample.number", "Sample"),
  colnames(sd)
)

sd_id_map <- sd |>
  dplyr::mutate(dplyr::across(dplyr::any_of(c("sample_name", "fecal.sample.number", "Sample")), as.character)) |>
  tidyr::pivot_longer(
    cols = dplyr::all_of(id_cols),
    names_to = "id_type",
    values_to = "join_id"
  ) |>
  dplyr::filter(!is.na(.data$join_id), .data$join_id != "") |>
  dplyr::distinct(.data$join_id, .keep_all = TRUE) |>
  dplyr::select(-.data$id_type)

removed_tbl <- tibble::tibble(
  sample_name = low_read_removed,
  reads_after_tax_filters = unname(reads_after_tax_filters[low_read_removed])
) |>
  # Try matching on multiple identifiers (rowname sample_name, fecal.sample.number, Sample),
  # and also allow "f###" ↔ "###" matching.
  dplyr::mutate(sample_id_stripped_f = sub("^f", "", .data$sample_name)) |>
  dplyr::left_join(sd_id_map, by = c("sample_name" = "join_id")) |>
  dplyr::left_join(
    sd_id_map,
    by = c("sample_id_stripped_f" = "join_id"),
    suffix = c("", "__stripped")
  ) |>
  dplyr::mutate(
    # If direct join failed, fill from stripped join columns when present.
    # Exclude columns that already end with __stripped: across would otherwise look for
    # e.g. Tank.ID__stripped__stripped when coalescing the suffixed duplicate.
    dplyr::across(
      .cols = -dplyr::any_of(c("sample_name", "sample_id_stripped_f", "reads_after_tax_filters")) &
        -dplyr::ends_with("__stripped"),
      .fns = \(x) dplyr::coalesce(x, get(paste0(dplyr::cur_column(), "__stripped")))
    )
  ) |>
  dplyr::select(-dplyr::ends_with("__stripped")) |>
  dplyr::arrange(.data$reads_after_tax_filters, .data$sample_name)

gt::gt(removed_tbl) |>
  gt::tab_header(
    title = "Samples removed for low read depth",
    subtitle = paste0("Computed on the taxa-filtered intermediate; threshold = ", min_sample_reads, " reads")
  ) |>
  gt::fmt_number(columns = "reads_after_tax_filters", decimals = 0) |>
  gt::cols_label(
    sample_name = "Sample",
    reads_after_tax_filters = "Reads (after taxa filters)"
  )
```

### Taxa removed by each taxa filter step (IDs only)

```{r}
taxa0 <- phyloseq::taxa_names(steps$ps0)
taxa1 <- phyloseq::taxa_names(steps$ps1)
taxa2 <- phyloseq::taxa_names(steps$ps2)
taxa3 <- phyloseq::taxa_names(steps$ps3)

tbl_taxa_removed <- dplyr::bind_rows(
  tibble::tibble(step = "step1_prevalence_taxa", taxon = setdiff(taxa0, taxa1)),
  tibble::tibble(step = "step2_taxonomy_taxa", taxon = setdiff(taxa1, taxa2)),
  tibble::tibble(step = "step3_post_sample_prune_zero_taxa", taxon = setdiff(taxa2, taxa3))
) |>
  dplyr::arrange(.data$step, .data$taxon)

tbl_taxa_removed |>
  dplyr::count(.data$step, name = "n_taxa_removed") |>
  gt::gt() |>
  gt::tab_header(
    title = "Taxa removed by preprocessing (counts)",
    subtitle = "Detailed ASV IDs are shown in a scrollable table below."
  ) |>
  gt::cols_label(step = "Filter step", n_taxa_removed = "Taxa removed"
  )

gt::gt(tbl_taxa_removed) |>
  gt::tab_header(
    title = "Taxa removed by preprocessing (ASV IDs)",
    subtitle = "Scrollable audit table (HTML)."
  ) |>
  gt::cols_label(step = "Filter step", taxon = "Taxon") |>
  gt::tab_options(container.height = gt::px(300), table.font.size = gt::px(12))
```

## Consistency check: does the replayed pipeline match the saved cleaned object?

The replayed pipeline *should* match the saved `path.pseq.cleaned` with respect to sample/taxa sets
after the same filters are applied (aside from optional decontam behavior and any subsequent metadata-based
filtering applied later in the pipeline).

```{r}
same_samples <- setequal(phyloseq::sample_names(ps_cleaned), phyloseq::sample_names(steps$ps3))
same_taxa <- setequal(phyloseq::taxa_names(ps_cleaned), phyloseq::taxa_names(steps$ps3))

tibble::tibble(
  check = c("Same sample set?", "Same taxa set?"),
  value = c(same_samples, same_taxa)
) |>
  gt::gt() |>
  gt::tab_header(title = "Replay vs saved cleaned object")
```

