---
title: "Experimental design overview"
subtitle: "00__Overview — summary of sequential stressor design"
author: "Michael Sieler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 10
    fig_height: 6
params:
  rerun_analysis: false
---

```{r setup, include=FALSE}
# Knit from any working directory: locate Sieler2026 repo root, then source_project_init.
knitr::opts_chunk$set(
  echo = isTRUE(params$rerun_analysis),
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

start <- if (requireNamespace("knitr", quietly = TRUE) && !is.null(knitr::current_input())) {
  dirname(knitr::current_input(dir = TRUE))
} else {
  getwd()
}
init_rel <- file.path("Code", "00__Setup", "source_project_init.R")
root <- normalizePath(start, mustWork = TRUE)
while (!file.exists(file.path(root, init_rel))) {
  parent <- dirname(root)
  if (identical(parent, root)) {
    stop("Could not find ", init_rel, " above ", start)
  }
  root <- parent
}
source(file.path(root, init_rel))
sieler2026_source_initialize_project(root = root)

# Knit with repo root as working directory so relative paths resolve consistently.
knitr::opts_knit$set(root.dir = root)

schematic_png <- here::here("Data", "Context", "ExperimentalDesignSchematic.png")

bundle_path <- here::here("Results", "00__Overview", "Stats", "overview__bundle.rds")

if (isTRUE(params$rerun_analysis)) {
  source(here::here("Code", "01__Analysis", "00__Overview.R"))
}

if (!file.exists(bundle_path)) {
  stop(
    "Bundle not found: ", bundle_path, "\n",
    "Run `Code/01__Analysis/00__Overview.R` from the project root, or knit this document with ",
    "`params$rerun_analysis: true` (requires full analysis environment)."
  )
}

overview <- readRDS(bundle_path)
```

```{r schematic, echo=FALSE, out.width="100%", fig.cap="Schematic of the sequential exposure experiment (antibiotics → temperature → parasite), with sampling endpoints."}
if (!file.exists(schematic_png)) {
  stop("Missing schematic PNG: ", schematic_png)
}
knitr::include_graphics(normalizePath(schematic_png, winslash = "/", mustWork = TRUE), rel_path = FALSE)
```

```{r sample-counts-gt, echo=FALSE}
overview$table_sample_counts
```

*Note:* This table is computed by `Code/01__Analysis/00__Overview.R` and loaded from `Results/00__Overview/Stats/overview__bundle.rds`. Values are **sample counts** (rows in `ps.list$All`), so counts at later timepoints can differ across exposure regimes due to mortality and downstream sample availability.

## Which samples are removed by the read-depth filter?

The cleaned phyloseq object (`path.pseq.cleaned`) is generated by `Code/00__Setup/04__DataPreProcess.R`,
which applies `apply_pre_analysis_filters()` including a minimum read-depth threshold after taxa filtering.
This chunk reports which samples fail that threshold (using only phyloseq objects stored inside this repo).

```{r low-read-removed-samples-gt, echo=FALSE}
# Created by: Michael Sieler
# Date last updated: 2026-04-22

ps_uncleaned_overview <- readRDS(path.pseq.uncleaned)
ps_cleaned_overview <- readRDS(path.pseq.cleaned)

# Match current preprocessing defaults (see Code/00__Setup/04__DataPreProcess.R)
min_prevalence_taxa_overview <- 3L
min_sample_reads_overview <- 5000L

# Recreate the taxa-filtered intermediate that precedes the low-read pruning step.
recreate_pre_low_read_ps <- function(ps_input, min_prevalence) {
  otu <- as(phyloseq::otu_table(ps_input), "matrix")
  if (!phyloseq::taxa_are_rows(ps_input)) {
    otu <- t(otu)
  }
  taxa_prev <- rowSums(otu > 0)
  ps_out <- phyloseq::prune_taxa(taxa_prev >= min_prevalence, ps_input)

  if (!is.null(phyloseq::tax_table(ps_out, errorIfNULL = FALSE))) {
    tax_df <- as.data.frame(as(phyloseq::tax_table(ps_out), "matrix"), stringsAsFactors = FALSE)
    rm_target <- apply(
      tax_df,
      1,
      function(x) any(grepl("Mitochondria|Chloroplast|Eukaryota", x, ignore.case = TRUE))
    )
    phylum_col <- if ("Phylum" %in% colnames(tax_df)) {
      "Phylum"
    } else if (ncol(tax_df) >= 2) {
      colnames(tax_df)[2]
    } else {
      NA_character_
    }
    unresolved <- rep(FALSE, nrow(tax_df))
    if (!is.na(phylum_col)) {
      phylum_vals <- tax_df[[phylum_col]]
      unresolved <- is.na(phylum_vals) |
        phylum_vals == "" |
        grepl("^Kingdom_Bacteria$", phylum_vals, ignore.case = TRUE) |
        grepl("^Bacteria Phylum$", phylum_vals, ignore.case = TRUE)
    }
    keep <- !(rm_target | unresolved)
    ps_out <- phyloseq::prune_taxa(rownames(tax_df)[keep], ps_out)
  }

  ps_out
}

ps_pre_low_read <- recreate_pre_low_read_ps(
  ps_uncleaned_overview,
  min_prevalence = min_prevalence_taxa_overview
)

reads_after_tax_filter <- phyloseq::sample_sums(ps_pre_low_read)
low_read_removed <- names(reads_after_tax_filter)[reads_after_tax_filter < min_sample_reads_overview]

# Add metadata columns where available, but keep this robust to missing fields.
sd <- phyloseq::sample_data(ps_uncleaned_overview) |>
  as.data.frame() |>
  tibble::rownames_to_column("sample_name") |>
  tibble::as_tibble()

removed_tbl <- tibble::tibble(
  sample_name = low_read_removed,
  reads_after_tax_filter = unname(reads_after_tax_filter[low_read_removed]),
  present_in_cleaned = low_read_removed %in% phyloseq::sample_names(ps_cleaned_overview)
) |>
  dplyr::left_join(sd, by = "sample_name") |>
  dplyr::arrange(reads_after_tax_filter, sample_name)

gt::gt(removed_tbl) |>
  gt::tab_header(
    title = "Samples removed for low read depth",
    subtitle = paste0(
      "Flagged after taxa filtering: reads < ", min_sample_reads_overview,
      " (min prevalence taxa = ", min_prevalence_taxa_overview, ")"
    )
  ) |>
  gt::fmt_number(columns = "reads_after_tax_filter", decimals = 0) |>
  gt::cols_label(
    sample_name = "Sample",
    reads_after_tax_filter = "Reads (after taxa filters)",
    present_in_cleaned = "Present in cleaned phyloseq?"
  )
```

## Study question (high-level)

This study tests whether **historical stress exposure** (prior perturbations) shapes how adult zebrafish and their gut microbiomes respond to *subsequent* challenges, with an emphasis on **resistance** (response during perturbation) and **resilience** (recovery afterward).

This document is intended to be rendered as HTML; the companion render script now lives at `Code/01__Analysis/00__Overview.R`.

## Model system and replication

- **Organism**: adult zebrafish (*Danio rerio*, 5D strain)
- **Total fish**: 720 individuals
- **Replication**: 24 tanks (replicated design)

## Treatment structure (sequential stressors)

Fish experienced up to three stressors applied **in sequence** across 60 days:

- **Antibiotics (A)**: days 0–14, delivered via medicated gelatin feed (streptomycin, ciprofloxacin, ampicillin).
- **Temperature (T)**: days 15–29, water raised from 28 °C → 35 °C, held, then returned to 28 °C.
- **Parasite (P)**: days 30–60, continuous exposure to *Pseudocapillaria tomentosa* eggs (flow-through effluent manifold).

Together these define a **fully factorial exposure history** with eight regimes: \(A^{\pm} \times T^{\pm} \times P^{\pm}\).

## Timeline (phases)

```{r timeline-gt}
timeline_tbl <- tibble::tibble(
  phase = c("1", "2", "3"),
  days = c("0–14", "15–29", "30–60"),
  stressor = c("Antibiotics", "Temperature (heat stress)", "Parasite exposure"),
  objective = c(
    "Initial microbiome perturbation",
    "Secondary abiotic stress following antibiotics",
    "Biotic challenge following one or more prior stressors"
  )
)

gt::gt(timeline_tbl) |>
  gt::tab_header(
    title = "Experimental phases",
    subtitle = "Three sequential exposures spanning 60 days"
  ) |>
  gt::cols_label(
    phase = "Phase",
    days = "Days",
    stressor = "Stressor",
    objective = "Objective"
  )
```

## Sampling endpoints

Sampling occurred at baseline and after each exposure block:

```{r sampling-gt}
sampling_tbl <- tibble::tibble(
  day = c("0", "14", "29", "60"),
  timepoint_label = c("T0", "T14", "T29", "T60"),
  aligned_to = c("Baseline", "Post-antibiotics", "Post-temperature", "Post-parasite / final"),
  samples_collected = c(
    "Fecal; gut",
    "Fecal; gut",
    "Fecal; gut",
    "Fecal; gut; infection + mortality endpoints"
  )
)

gt::gt(sampling_tbl) |>
  gt::tab_header(
    title = "Sampling schedule",
    subtitle = "Primary collection points used across microbiome and host outcomes"
  ) |>
  gt::cols_label(
    day = "Day",
    timepoint_label = "Timepoint",
    aligned_to = "Interpretation",
    samples_collected = "Primary samples / endpoints"
  )
```

## Primary outcome domains

- **Host outcomes**
  - Mortality
  - Infection prevalence and worm burden (day 60 endpoint)
- **Gut microbiome (16S rRNA)**
  - Alpha diversity and beta diversity
  - Community compositional displacement and recovery
  - Taxon-specific abundance changes across exposure histories
- **Host transcriptomics (RNA-seq)**
  - Intestinal gene expression at the day 60 endpoint (subset of fish)

## Key design features (interpretation)

- **Sequential (not simultaneous) exposures** emulate realistic cumulative stress histories.
- Mixes **abiotic** (antibiotics, heat) and **biotic** (parasite) perturbations.
- The factorial exposure histories allow testing for:
  - main effects of each stressor history
  - interaction effects (e.g., antibiotic history × parasite outcomes)
  - evidence for historical contingency in both host and microbiome responses
