---
title: "Gut microbial community composition (display)"
subtitle: "02__Composition — mirrors `Code/01__Analysis/02__Composition.R`"
author: "Michael Sieler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 10
    fig_height: 6
params:
  rerun_analysis: false
---

```{r setup, include=FALSE}
# Knit from any working directory: locate Sieler2026 repo root, then source_project_init.
knitr::opts_chunk$set(
  echo = isTRUE(params$rerun_analysis),
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

start <- if (requireNamespace("knitr", quietly = TRUE) && !is.null(knitr::current_input())) {
  dirname(knitr::current_input(dir = TRUE))
} else {
  getwd()
}
init_rel <- file.path("Code", "00__Setup", "source_project_init.R")
root <- normalizePath(start, mustWork = TRUE)
while (!file.exists(file.path(root, init_rel))) {
  parent <- dirname(root)
  if (identical(parent, root)) {
    stop("Could not find ", init_rel, " above ", start)
  }
  root <- parent
}
source(file.path(root, init_rel))
sieler2026_source_initialize_project(root = root)

# Knit with repo root as working directory so relative paths resolve consistently.
knitr::opts_knit$set(root.dir = root)

bundle_path <- here::here("Results", "02__Composition", "Stats", "composition__gut__bundle.rds")

if (isTRUE(params$rerun_analysis)) {
  source(here::here("Code", "01__Analysis", "02__Composition.R"))
}

if (!file.exists(bundle_path)) {
  stop(
    "Bundle not found: ", bundle_path, "\n",
    "Run `Code/01__Analysis/02__Composition.R` from the project root, or knit this document with ",
    "`params$rerun_analysis: true` (requires full analysis environment)."
  )
}

comp <- readRDS(bundle_path)

# Resolve paths saved in bundles:
# - Prefer the stored path when it exists on this machine
# - Otherwise, treat it as relative to the repo root (here::here)
resolve_bundle_path <- function(p) {
  if (is.null(p) || !nzchar(p)) {
    return(NA_character_)
  }
  if (file.exists(p)) {
    return(p)
  }
  p2 <- tryCatch(here::here(p), error = function(e) NA_character_)
  if (!is.na(p2) && file.exists(p2)) {
    return(p2)
  }
  p3 <- file.path(root, p)
  if (file.exists(p3)) {
    return(p3)
  }
  p
}

# Include a PNG from the bundle, robust to absolute/relative paths.
# We rely on Pandoc `--embed-resources` (RStudio knit) to create a standalone HTML.
include_png <- function(p) {
  p_res <- resolve_bundle_path(p)
  if (is.null(p_res) || !nzchar(p_res)) {
    return(invisible(NULL))
  }
  p_abs <- normalizePath(p_res, winslash = "/", mustWork = TRUE)
  knitr::include_graphics(p_abs, rel_path = FALSE)
}

# Helper: safely fetch a module PNG path (returns NA if missing).
comp_png <- function(module, sub) {
  mod <- comp$modules[[module]]
  if (is.null(mod) || is.null(mod[[sub]])) {
    return(NA_character_)
  }
  p <- mod[[sub]]$paths[["figure_png"]]
  p <- resolve_bundle_path(p)
  if (is.na(p) || !nzchar(p) || !file.exists(p)) {
    NA_character_
  } else {
    p
  }
}
```

This report loads `Results/02__Composition/Stats/composition__gut__bundle.rds` and displays **gt tables** (PERMANOVA and betadisper ANOVA; Bray–Curtis and Canberra) and **figures** from [`Code/01__Analysis/02__Composition.R`](../01__Analysis/02__Composition.R). The driver is not re-run unless you knit with `params$rerun_analysis: true`.

---

## Statistical framework

```{r model-text, echo=FALSE, results='asis'}
if (!is.null(comp$meta$description)) {
  cat(comp$meta$description, "\n\n")
}
if (!is.null(comp$meta$model_table_c)) {
  cat("**Table C (PERMANOVA, stress history):**\n\n")
  cat(comp$meta$model_table_c, "\n\n")
}
if (!is.null(comp$meta$model_table_d)) {
  cat("**Table D (PERMANOVA, stress history × parasite — interaction model):**\n\n")
  cat(comp$meta$model_table_d, "\n\n")
}
if (!is.null(comp$meta$model_table_dispersion_history)) {
  cat("**Dispersion — stress history (betadisper ANOVA):**\n\n")
  cat(comp$meta$model_table_dispersion_history, "\n\n")
}
if (!is.null(comp$meta$model_table_dispersion_parasite)) {
  cat("**Dispersion — stress history × parasite (betadisper ANOVA):**\n\n")
  cat(comp$meta$model_table_dispersion_parasite, "\n\n")
}
if (!is.null(comp$meta$model_table_parasite_within_permanova)) {
  cat("**Table E (PERMANOVA, stratified — Parasite within each history level):**\n\n")
  cat(comp$meta$model_table_parasite_within_permanova, "\n\n")
}
if (!is.null(comp$meta$model_table_parasite_within_betadisper)) {
  cat("**Table F (betadisper ANOVA, stratified — Parasite within each history level):**\n\n")
  cat(comp$meta$model_table_parasite_within_betadisper, "\n\n")
}
# Backward compatibility if an older bundle only had diversity-style meta keys
if (!is.null(comp$meta$model_linear_trend)) {
  cat("**Linear trend (legacy meta key):**\n\n")
  cat(comp$meta$model_linear_trend, "\n\n")
}
if (!is.null(comp$meta$model_table2)) {
  cat("**P-value summary (legacy meta key):**\n\n")
  cat(comp$meta$model_table2, "\n\n")
}
if (!is.null(comp$meta$model_interaction)) {
  cat("**History × parasite (legacy meta key):**\n\n")
  cat(comp$meta$model_interaction, "\n\n")
}
if (!is.null(comp$meta$model_formulas)) {
  gt::gt(tibble::enframe(comp$meta$model_formulas, name = "Model", value = "Formula")) %>%
    gt::tab_header(title = "Model formulas (reference)")
}
if (!is.null(comp$meta$n_perm)) {
  cat(
    "\n\n*Permutations:* ", comp$meta$n_perm,
    if (!is.null(comp$meta$n_processes)) {
      paste0("; parallel processes: ", comp$meta$n_processes)
    } else {
      ""
    },
    "\n\n"
  )
}
```

---

## Tables

## Exposure regimes (factorial A × T × P)

### PERMANOVA: Antibiotics × Temperature × Parasite (Bray–Curtis and Canberra)

```{r table-permanova-factorial-atp, echo=FALSE}
if (!is.null(comp$table_permanova_factorial_atp)) {
  comp$table_permanova_factorial_atp
}
```

### Betadisper ANOVA: factorial exposure regimes (Bray–Curtis and Canberra)

```{r table-bdisp-factorial-atp, echo=FALSE}
if (!is.null(comp$table_betadisper_anova_factorial_atp)) {
  comp$table_betadisper_anova_factorial_atp
}
```

### Table C — PERMANOVA: stress history (Bray–Curtis and Canberra)

```{r table-permanova-history, echo=FALSE}
if (!is.null(comp$table_permanova_stress_history)) {
  comp$table_permanova_stress_history
}
```

### Table D — PERMANOVA: stress history × parasite (interaction model)

```{r table-permanova-hp, echo=FALSE}
if (!is.null(comp$table_permanova_stress_history_parasite)) {
  comp$table_permanova_stress_history_parasite
}
```

### Betadisper ANOVA — stress history (Bray–Curtis and Canberra)

```{r table-bdisp-anova-hist, echo=FALSE}
if (!is.null(comp$table_betadisper_anova_stress_history)) {
  comp$table_betadisper_anova_stress_history
}
```

### Betadisper ANOVA — stress history × parasite (Bray–Curtis and Canberra)

```{r table-bdisp-anova-hp, echo=FALSE}
if (!is.null(comp$table_betadisper_anova_stress_history_parasite)) {
  comp$table_betadisper_anova_stress_history_parasite
}
```

### Table E — PERMANOVA: Parasite within each prior stressor history (stratified)

```{r table-permanova-parasite-within, echo=FALSE}
if (!is.null(comp$table_permanova_parasite_within_history)) {
  comp$table_permanova_parasite_within_history
}
```

### Table F — Betadisper ANOVA: Parasite within each prior stressor history (stratified)

```{r table-bdisp-parasite-within, echo=FALSE}
if (!is.null(comp$table_betadisper_parasite_within_history)) {
  comp$table_betadisper_parasite_within_history
}
```

### Legacy combined slots (diversity-style, usually `NULL` for composition)

```{r table-legacy-combined, echo=FALSE, results='asis'}
if (!is.null(comp$table_combined_trends)) {
  cat("### Linear stress-history trends\n\n")
  print(comp$table_combined_trends)
}
if (!is.null(comp$table_combined_interaction)) {
  cat("\n\n### History × parasite interaction\n\n")
  print(comp$table_combined_interaction)
}
if (!is.null(comp$table_combined_table2)) {
  cat("\n\n### P-value summary (combined table 2)\n\n")
  print(comp$table_combined_table2)
}
```

### Other `table_*` objects (forward compatibility)

```{r table-extra, echo=FALSE, results='asis'}
reserved <- c(
  "table_combined_trends", "table_combined_interaction", "table_combined_table2",
  "table_permanova_stress_history", "table_permanova_stress_history_parasite",
  "table_betadisper_anova_stress_history", "table_betadisper_anova_stress_history_parasite",
  "table_permanova_parasite_within_history", "table_betadisper_parasite_within_history",
  "tables_tidy", "modules", "meta"
)
extra <- setdiff(names(comp), reserved)
extra <- extra[grepl("^table_", extra)]
if (length(extra) > 0L) {
  for (nm in extra) {
    cat("\n### ", nm, "\n\n")
    print(comp[[nm]])
  }
}
```

---

## Figures

Files live under `Results/02__Composition/Figures/` (PDF and PNG). Below uses **PNG** via `knitr::include_graphics()` for reliable HTML embedding. ggplot objects are also stored in `comp$modules` if you want to tweak layers in an interactive session.

### Relative abundance by exposure regime (genus, merged samples)

```{r fig-rel-abund, echo=FALSE, fig.width=8, fig.height=8, fig.cap="Relative abundance by exposure regime (genus level), faceted by prior stressor history; top 8 genera plus Other. Samples merged by Treatment."}
p <- comp_png("relative_abundance", "treatment")
if (!is.na(p)) {
  include_png(p)
}
```

### PCoA — factorial exposure regimes (Bray–Curtis)

```{r fig-pcoa-factorial-atp-bray, echo=FALSE, fig.width=10, fig.height=7, fig.cap="PCoA of Bray–Curtis distances faceted by Antibiotics and Temperature; points colored by parasite exposure."}
p <- comp$modules$beta_bray$factorial_atp$pcoa$paths[["figure_png"]]
if (!is.null(p)) {
  include_png(p)
}
```

### Betadisper — factorial exposure regimes (Bray–Curtis)

```{r fig-bdisp-factorial-atp-bray, echo=FALSE, fig.width=10, fig.height=7, fig.cap="Betadisper distances by Antibiotics × Temperature × Parasite groups; Bray–Curtis."}
p <- comp$modules$beta_bray$factorial_atp$betadisper$paths[["figure_png"]]
if (!is.null(p)) {
  include_png(p)
}
```

### PCoA — factorial exposure regimes (Canberra)

```{r fig-pcoa-factorial-atp-canberra, echo=FALSE, fig.width=10, fig.height=7, fig.cap="PCoA of Canberra distances faceted by Antibiotics and Temperature; points colored by parasite exposure."}
p <- comp$modules$beta_canberra$factorial_atp$pcoa$paths[["figure_png"]]
if (!is.null(p)) {
  include_png(p)
}
```

### Betadisper — factorial exposure regimes (Canberra)

```{r fig-bdisp-factorial-atp-canberra, echo=FALSE, fig.width=10, fig.height=7, fig.cap="Betadisper distances by Antibiotics × Temperature × Parasite groups; Canberra."}
p <- comp$modules$beta_canberra$factorial_atp$betadisper$paths[["figure_png"]]
if (!is.null(p)) {
  include_png(p)
}
```

### PCoA — stressor history (Bray–Curtis)

```{r fig-pcoa-history, echo=FALSE, fig.cap="Unconstrained PCoA of Bray–Curtis distances; points and ellipses colored by stressor history."}
p <- comp_png("beta_bray", "pcoa_history")
if (!is.na(p)) {
  include_png(p)
}
```

### PCoA — stressor history and parasite exposure (Bray–Curtis)

```{r fig-pcoa-hp, echo=FALSE, fig.width=9, fig.height=10, fig.cap="PCoA with parasite exposure indicated by point fill and ellipse linetype."}
p <- comp_png("beta_bray", "pcoa_history_parasite")
if (!is.na(p)) {
  include_png(p)
}
```

### Betadisper — stressor history (Bray–Curtis)

```{r fig-bdisp-history, echo=FALSE, fig.width=8, fig.height=8, fig.cap="Distance to group centroid (betadisper) by stressor history."}
p <- comp_png("beta_bray", "betadisper_history")
if (!is.na(p)) {
  include_png(p)
}
```

### Betadisper — stressor history × parasite (Bray–Curtis)

```{r fig-bdisp-hp, echo=FALSE, fig.width=9, fig.height=10, fig.cap="Betadisper distances for HistoryLevel × Parasite groups (six levels)."}
p <- comp_png("beta_bray", "betadisper_history_parasite")
if (!is.na(p)) {
  include_png(p)
}
```

### PCoA — stressor history (Canberra)

```{r fig-pcoa-canberra-history, echo=FALSE, fig.cap="Unconstrained PCoA of Canberra distances; points and ellipses colored by stressor history."}
p <- comp_png("beta_canberra", "pcoa_history")
if (!is.na(p)) {
  include_png(p)
}
```

### PCoA — stressor history and parasite exposure (Canberra)

```{r fig-pcoa-canberra-hp, echo=FALSE, fig.width=9, fig.height=10, fig.cap="PCoA on Canberra distances with parasite exposure indicated by point fill and ellipse linetype."}
p <- comp_png("beta_canberra", "pcoa_history_parasite")
if (!is.na(p)) {
  include_png(p)
}
```

### Betadisper — stressor history (Canberra)

```{r fig-bdisp-canberra-history, echo=FALSE, fig.cap="Distance to group centroid (betadisper) by stressor history; Canberra metric."}
p <- comp_png("beta_canberra", "betadisper_history")
if (!is.na(p)) {
  include_png(p)
}
```

### Betadisper — stressor history × parasite (Canberra)

```{r fig-bdisp-canberra-hp, echo=FALSE, fig.width=9, fig.height=10, fig.cap="Betadisper distances (Canberra) for HistoryLevel × Parasite groups."}
p <- comp_png("beta_canberra", "betadisper_history_parasite")
if (!is.na(p)) {
  include_png(p)
}
```

### PCoA — parasite exposure faceted by prior stressor history (Bray–Curtis)

```{r fig-pcoa-bray-faceted, echo=FALSE, fig.width=12, fig.height=5, fig.cap="Global PCoA (cmdscale on Bray–Curtis); one ordination for all samples, faceted by prior stressor history; points colored by parasite exposure."}
p <- comp_png("beta_bray", "pcoa_parasite_faceted_history")
if (!is.na(p)) {
  include_png(p)
}
```

### PCoA — parasite exposure faceted by prior stressor history (Canberra)

```{r fig-pcoa-canberra-faceted, echo=FALSE, fig.width=12, fig.height=5, fig.cap="Global PCoA (cmdscale on Canberra); same layout as the Bray–Curtis faceted figure."}
p <- comp_png("beta_canberra", "pcoa_parasite_faceted_history")
if (!is.na(p)) {
  include_png(p)
}
```

### PCoA — exposure regime (points) with parasite ellipses (Bray–Curtis)

Same global ordination as the figures above; **points** use the eight exposure regimes (`Treatment`: A/T/P combinations); **ellipses** remain grouped by parasite exposure.

```{r fig-pcoa-bray-faceted-regime-points, echo=FALSE, fig.width=12, fig.height=6.5, fig.cap="Global PCoA (cmdscale on Bray–Curtis); points colored by exposure regime; ellipses by parasite exposure within each prior stressor history facet."}
p <- comp_png("beta_bray", "pcoa_parasite_faceted_history_regime_points")
if (!is.na(p)) {
  include_png(p)
}
```

### PCoA — exposure regime (points) with parasite ellipses (Canberra)

```{r fig-pcoa-canberra-faceted-regime-points, echo=FALSE, fig.width=12, fig.height=6.5, fig.cap="Global PCoA (cmdscale on Canberra); points colored by exposure regime; ellipses by parasite exposure within each facet."}
p <- comp_png("beta_canberra", "pcoa_parasite_faceted_history_regime_points")
if (!is.na(p)) {
  include_png(p)
}
```

---

## Bundle metadata

```{r meta, echo=FALSE}
meta_simple <- comp$meta[
  intersect(
    names(comp$meta),
    c(
      "run_date", "ps_list_element", "bundle_version", "script", "description",
      "model_linear_trend", "model_table2", "model_interaction",
      "model_table_c", "model_table_d", "model_table_dispersion_history", "model_table_dispersion_parasite",
      "model_table_parasite_within_permanova", "model_table_parasite_within_betadisper",
      "n_perm", "n_processes"
    )
  )
]
gt::gt(tibble::enframe(meta_simple, name = "Field", value = "Value")) %>%
  gt::tab_header(title = "Bundle provenance")
if (!is.null(comp$meta$model_formulas)) {
  gt::gt(tibble::enframe(comp$meta$model_formulas, name = "Model", value = "Formula")) %>%
    gt::tab_header(title = "Model formulas (reference)")
}
```

