---
title: "Gut microbial differential abundance — MaAsLin3 (display)"
subtitle: "03__DiffAbund — mirrors `Code/01__Analysis/03__DiffAbund.R`"
author: "Michael Sieler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 10
    fig_height: 6
params:
  rerun_analysis: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = isTRUE(params$rerun_analysis),
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)

start <- if (requireNamespace("knitr", quietly = TRUE) && !is.null(knitr::current_input())) {
  dirname(knitr::current_input(dir = TRUE))
} else {
  getwd()
}
init_rel <- file.path("Code", "00__Setup", "source_project_init.R")
root <- normalizePath(start, mustWork = TRUE)
while (!file.exists(file.path(root, init_rel))) {
  parent <- dirname(root)
  if (identical(parent, root)) {
    stop("Could not find ", init_rel, " above ", start)
  }
  root <- parent
}
source(file.path(root, init_rel))
sieler2026_source_initialize_project(root = root)

bundle_path <- here::here("Results", "03__DiffAbund", "Stats", "diffabund__gut__bundle.rds")

if (isTRUE(params$rerun_analysis)) {
  source(here::here("Code", "01__Analysis", "03__DiffAbund.R"))
}

if (!file.exists(bundle_path)) {
  stop(
    "Bundle not found: ", bundle_path, "\n",
    "Run `Code/01__Analysis/03__DiffAbund.R` from the project root, or knit with ",
    "`params$rerun_analysis: true` (requires `ps.list`, `maaslin3`, and full analysis environment)."
  )
}

dab <- readRDS(bundle_path)
```

This report loads `Results/03__DiffAbund/Stats/diffabund__gut__bundle.rds` from [`Code/01__Analysis/03__DiffAbund.R`](../01__Analysis/03__DiffAbund.R). To re-run MaAsLin3 from R Markdown, knit with `rerun_analysis: true`.


```{r meta, echo=FALSE}
dab$meta
```

---

## Significant taxa counts by exposure regime (no-tank)

```{r sig-counts-table, echo=FALSE}
if (!is.null(dab$table_sig_counts_by_exposure_regime_noTank)) {
  dab$table_sig_counts_by_exposure_regime_noTank
} else {
  cat("Table not found in bundle. Re-run `Code/01__Analysis/03__DiffAbund.R` to regenerate outputs.")
}
```

---

## Significant taxa counts by prior stressor history (tank)

```{r sig-counts-stress-history, echo=FALSE}
if (!is.null(dab$table_sig_counts_by_stress_history_tank)) {
  dab$table_sig_counts_by_stress_history_tank
} else {
  cat("Table not found in bundle. Re-run `Code/01__Analysis/03__DiffAbund.R` to regenerate outputs.")
}
```

## Significant taxa counts by stress history × parasite exposure (tank)

```{r sig-counts-stress-history-parasite, echo=FALSE}
if (!is.null(dab$table_sig_counts_by_stress_history_x_parasite_tank)) {
  dab$table_sig_counts_by_stress_history_x_parasite_tank
} else {
  cat("Table not found in bundle. Re-run `Code/01__Analysis/03__DiffAbund.R` to regenerate outputs.")
}
```

## Mortality: significant taxa vs percent_mortality (tank effect model)

```{r mortality-top-table, echo=FALSE}
if (!is.null(dab$table_top10_taxa_associated_with_mortality_tank)) {
  dab$table_top10_taxa_associated_with_mortality_tank
} else {
  cat("Table not found in bundle. Re-run `Code/01__Analysis/03__DiffAbund.R` to regenerate outputs.")
}
```

## Figures: mortality-associated taxa (exported from driver)

```{r mortality-figs, echo=FALSE, fig.width=12, fig.height=8, out.width="100%"}
figs <- dab$figures

paths <- c(
  figs$mortality_top_taxa_coef_bar_tank,
  figs$mortality_top_taxa_scatter_facets_tank,
  figs$mortality_top_taxa_scatter_combined_tank,
  figs$mortality_culicoidibacter_scatter_tank
)

paths <- paths[!is.na(paths) & nzchar(paths)]
paths <- paths[file.exists(paths)]

if (length(paths) > 0L) {
  knitr::include_graphics(paths)
} else {
  cat(
    "Mortality figures not found in bundle outputs. Re-run `Code/01__Analysis/03__DiffAbund.R` ",
    "to export figures to `Results/03__DiffAbund/Figures/`."
  )
}
```

## MaAsLin output paths

```{r paths, echo=FALSE}
dab$paths
```
