---
title: "Mortality and infection vs prior stressor history (display)"
author: "Michael Sieler"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 6
    fig_height: 6
params:
  rerun_analysis: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = isTRUE(params$rerun_analysis),
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 6,
  dpi = 300
)

start <- if (requireNamespace("knitr", quietly = TRUE) && !is.null(knitr::current_input())) {
  dirname(knitr::current_input(dir = TRUE))
} else {
  getwd()
}
init_rel <- file.path("Code", "00__Setup", "source_project_init.R")
root <- normalizePath(start, mustWork = TRUE)
while (!file.exists(file.path(root, init_rel))) {
  parent <- dirname(root)
  if (identical(parent, root)) {
    stop("Could not find ", init_rel, " above ", start)
  }
  root <- parent
}
source(file.path(root, init_rel))
sieler2026_source_initialize_project(root = root)

bundle_path <- here::here("Results", "05__Mort-Inf", "Stats", "mortinf__host__bundle.rds")

if (isTRUE(params$rerun_analysis)) {
  source(here::here("Code", "01__Analysis", "05__Mort-Inf.R"))
}

if (!file.exists(bundle_path)) {
  stop(
    "Bundle not found: ", bundle_path, "\n",
    "Run `Code/01__Analysis/05__Mort-Inf.R` from the project root, or knit with ",
    "`params$rerun_analysis: true`."
  )
}

mortinf <- readRDS(bundle_path)

# Resolve paths saved in bundles:
# - Prefer the stored path when it exists on this machine
# - Otherwise, treat it as relative to the repo root (here::here)
resolve_bundle_path <- function(p) {
  if (is.null(p) || !nzchar(p)) {
    return(NA_character_)
  }
  if (file.exists(p)) {
    return(p)
  }
  p2 <- tryCatch(here::here(p), error = function(e) NA_character_)
  if (!is.na(p2) && file.exists(p2)) {
    return(p2)
  }
  p3 <- file.path(root, p)
  if (file.exists(p3)) {
    return(p3)
  }
  p
}

# Include a PNG from the bundle, robust to absolute/relative paths.
include_png <- function(p) {
  p_res <- resolve_bundle_path(p)
  if (is.null(p_res) || !nzchar(p_res)) {
    return(invisible(NULL))
  }
  p_abs <- normalizePath(p_res, winslash = "/", mustWork = TRUE)
  knitr::include_graphics(p_abs, rel_path = FALSE)
}
```

This report loads `Results/05__Mort-Inf/Stats/mortinf__host__bundle.rds` produced by [`Code/01__Analysis/05__Mort-Inf.R`](../01__Analysis/05__Mort-Inf.R). **Mortality** uses binomial **glmmTMB** models: **HistoryLevelNum + (1 | Tank.ID)** for the main cumulative-stress trend figure, and **HistoryLevelNum × Parasite + (1 | Tank.ID)** for the supplementary parasite comparison. **Infection** is **Day-60 surviving P+ fish** (one row per `Sample`); tank-level models include **HistoryLevelNum** and **Treatment** with **emmeans** contrasts. The **four** manuscript figures below follow a 2×2 layout (bars on top, trends on bottom), **square 6 × 6 in** PNGs.

## Metadata

```{r meta, echo=FALSE}
mortinf$meta
```

***

## Main figures (manuscript, square)

### 1. Final mortality by exposure regime

```{r fig-mortality-bars-sq, echo=FALSE, fig.cap="Day-60 percent mortality by exposure regime (45 fish at risk per regime); facets = prior stressor history. Labels: percent and deaths/at-risk."}
include_png(mortinf$paths$figures$mortality_percent_by_exposure_regime_prior_history_square_png)
```

### 1b. Final mortality by exposure regime (A×T only; parasite pooled within stratum)

```{r fig-mortality-at-bars-pooled, echo=FALSE, fig.cap="Day-60 percent mortality by A×T exposure regime (parasite pooled); facets = prior stressor history. Labels: percent and deaths/at-risk pooled across P- and P+ tanks within each stratum."}
include_png(mortinf$paths$figures$mortality_percent_by_AT_regime_prior_history_pooled_parasite_png)
```

### 1c. Final mortality by exposure regime (A×T), faceted by stress history with P− vs P+ side-by-side

```{r fig-mortality-at-bars-by-parasite, echo=FALSE, fig.cap="Day-60 percent mortality by A×T exposure regime with P− vs P+ side-by-side; facets = prior stressor history. Outline color = stress history palette (green/orange/purple); fill = parasite exposure (grey/red)."}
include_png(mortinf$paths$figures$mortality_percent_by_AT_regime_prior_history_by_parasite_png)
```

### 2. Infection prevalence by exposure regime (P+ survivors)

```{r fig-infection-bars-sq, echo=FALSE, fig.cap="Infection prevalence in P+ regimes among Day-60 surviving sampled fish; facets = prior stressor history. Prevalence = worms present (Total.Worm.Count > 0) / survivors sampled; labels show percent and infected/total."}
include_png(mortinf$paths$figures$infection_prevalence_by_exposure_regime_prior_history_square_png)
```

### 3. Mortality trend: prior stressor history (main model)

```{r fig-mortality-trend-sq, echo=FALSE, fig.cap="Day-60 tank-level mortality: observed (gray) and GLMM predicted probabilities with 95% CI (black line, blue ribbon). Model: HistoryLevelNum + (1 | Tank.ID)."}
include_png(mortinf$paths$figures$mortality_prior_stressor_trend_square_png)
```

### 4. Infection prevalence trend (P+ survivors, Day 60)

```{r fig-infection-trend-sq, echo=FALSE, fig.cap="Tank-level infection prevalence among Day-60 surviving P+ fish; observed (gray) and GLMM predicted probabilities with 95% CI. Primary model: HistoryLevelNum + (1 | Tank.ID)."}
include_png(mortinf$paths$figures$infection_prevalence_trend_predicted_square_png)
```

***

## Supplementary figures: infection prevalence and worm burden (bar charts)

Descriptive bar plots mirror the exposure-regime manuscript style: **by prior stressor history** (three levels: 0, 1, or 2 prior stressors before the parasite phase) use the history color palette from the legacy experiment summary; **by exposure regime** panels repeat **total** and **mean (infected-only)** burden with facets for prior stressor history (Zero / One / Two), matching the infection prevalence faceted figure.

### Infection prevalence and burden by prior stressor history

```{r fig-supp-inf-prev-hist, echo=FALSE, fig.cap="Infection prevalence among Day-60 sampled P+ survivors, pooled by prior stressor history (0–2). Labels: percent and infected/total sampled."}
include_png(mortinf$paths$figures$infection_prevalence_by_prior_stressor_history_bars_square_png)
```

```{r fig-supp-inf-total-hist, echo=FALSE, fig.cap="Total worm counts (sum of Total.Worm.Count) across all sampled P+ survivors within each prior stressor history level."}
include_png(mortinf$paths$figures$infection_total_worm_count_by_prior_stressor_history_bars_square_png)
```

```{r fig-supp-inf-mean-hist, echo=FALSE, fig.cap="Mean worm burden among infected fish only (Total.Worm.Count > 0) by prior stressor history; n = infected fish in each level."}
include_png(mortinf$paths$figures$infection_mean_worm_burden_infected_by_prior_stressor_history_bars_square_png)
```

### Total and mean worm burden by exposure regime (faceted by prior stressor history)

```{r fig-supp-inf-total-regime, echo=FALSE, fig.cap="Total worm counts per exposure regime within each prior-stressor-history facet (same structure as infection prevalence by regime)."}
include_png(mortinf$paths$figures$infection_total_worm_count_by_exposure_regime_prior_history_square_png)
```

```{r fig-supp-inf-mean-regime, echo=FALSE, fig.cap="Mean worm burden among infected fish per exposure regime, faceted by prior stressor history."}
include_png(mortinf$paths$figures$infection_mean_worm_burden_infected_by_exposure_regime_prior_history_square_png)
```

***

## Supplementary figure: mortality P+ vs P− (interaction model)

```{r fig-mortality-parasite-sq, echo=FALSE, fig.cap="Predicted mortality by prior stressors and parasite challenge (P+ vs P-). Model: HistoryLevelNum * Parasite + (1 | Tank.ID)."}
include_png(mortinf$paths$figures$mortality_parasite_trend_comparison_square_png)
```

***

## Table: deaths and at-risk by exposure regime

```{r tbl-deaths, echo=FALSE}
tbl_deaths <- mortinf$tables$deaths_by_exposure_regime

gt_deaths <- tbl_deaths %>%
  gt::gt() %>%
  gt::tab_header(
    title = "Deaths and survivors by exposure regime (Day 60)",
    subtitle = "Per regime: 3 tanks × 15 fish = 45 at risk"
  ) %>%
  gt::cols_label(
    Treatment = "Exposure regime",
    deaths = "Deaths",
    at_risk = "At risk",
    survived = "Survived"
  ) %>%
  gt_bold_column_labels()

if (exists("treatment_color_scale", inherits = TRUE) && exists("treatment_order", inherits = TRUE)) {
  for (trt in treatment_order) {
    if (trt %in% names(treatment_color_scale)) {
      trt_color <- treatment_color_scale[[trt]]
      gt_deaths <- gt_deaths %>%
        gt::tab_style(
          style = gt::cell_fill(color = trt_color),
          locations = gt::cells_body(columns = "Treatment", rows = Treatment == trt)
        ) %>%
        gt::tab_style(
          style = gt::cell_text(color = "white", weight = "bold"),
          locations = gt::cells_body(columns = "Treatment", rows = Treatment == trt)
        )
    }
  }
}

gt_deaths
```

***

## Table: mortality (%) by exposure regime

```{r tbl-mort-pct, echo=FALSE}
tbl_pct <- mortinf$tables$mortality_percent_by_exposure_regime

gt_pct <- tbl_pct %>%
  gt::gt() %>%
  gt::tab_header(
    title = "Mortality percentage by exposure regime",
    subtitle = "Day 60; denominator 45 fish per regime"
  ) %>%
  gt::cols_label(
    Treatment = "Exposure regime",
    mortality_percent = "Mortality (%)"
  ) %>%
  gt_bold_column_labels()

if (exists("treatment_color_scale", inherits = TRUE) && exists("treatment_order", inherits = TRUE)) {
  for (trt in treatment_order) {
    if (trt %in% names(treatment_color_scale)) {
      trt_color <- treatment_color_scale[[trt]]
      gt_pct <- gt_pct %>%
        gt::tab_style(
          style = gt::cell_fill(color = trt_color),
          locations = gt::cells_body(columns = "Treatment", rows = Treatment == trt)
        ) %>%
        gt::tab_style(
          style = gt::cell_text(color = "white", weight = "bold"),
          locations = gt::cells_body(columns = "Treatment", rows = Treatment == trt)
        )
    }
  }
}

gt_pct
```

***

## Table: combined mortality summary (optional manuscript layout)

```{r tbl-mort-summary, echo=FALSE}
tbl_sum <- mortinf$tables$mortality_summary_by_exposure_regime %>%
  dplyr::transmute(
    Treatment = as.character(.data$Treatment),
    `At risk` = .data$at_risk,
    Survived = .data$alive,
    Deaths = .data$dead,
    `Mortality (%)` = .data$percent_mortality
  )

gt_sum <- tbl_sum %>%
  gt::gt() %>%
  gt::tab_header(
    title = "Mortality summary by exposure regime",
    subtitle = "Treatment, at risk, survived, deaths, and percent mortality (Day 60)"
  ) %>%
  gt_bold_column_labels()

if (exists("treatment_color_scale", inherits = TRUE) && exists("treatment_order", inherits = TRUE)) {
  for (trt in treatment_order) {
    if (trt %in% names(treatment_color_scale)) {
      trt_color <- treatment_color_scale[[trt]]
      gt_sum <- gt_sum %>%
        gt::tab_style(
          style = gt::cell_fill(color = trt_color),
          locations = gt::cells_body(columns = "Treatment", rows = Treatment == trt)
        ) %>%
        gt::tab_style(
          style = gt::cell_text(color = "white", weight = "bold"),
          locations = gt::cells_body(columns = "Treatment", rows = Treatment == trt)
        )
    }
  }
}

gt_sum
```

***

## Table: mortality by exposure regime and prior stressor history (descriptive)

```{r tbl-mort-regime-prior-history, echo=FALSE}
tbl_cross <- mortinf$tables$mortality_percent_by_exposure_regime_prior_history
tbl_cross <- tbl_cross %>%
  dplyr::transmute(
    `Prior stressor history` = as.character(.data$Prior_Stressor_History),
    `Exposure regime` = as.character(.data$Treatment),
    `At risk` = .data$at_risk,
    Survived = .data$alive,
    Deaths = .data$dead,
    `Mortality (%)` = .data$percent_mortality
  ) %>%
  dplyr::arrange(.data$`Prior stressor history`, .data$`Exposure regime`)

tbl_cross %>%
  gt::gt() %>%
  gt::tab_header(
    title = "Mortality by exposure regime and prior stressor history",
    subtitle = "Day 60; descriptive counts (same structure as faceted bar figure). 45 fish at risk per regime."
  ) %>%
  gt_fmt_numbers(tbl_cross, decimals = 1L) %>%
  gt_bold_column_labels()
```

***

## Table: mortality GLMM — marginal means (response scale)

```{r tbl-mort-glmm-emmeans, echo=FALSE}
mme <- mortinf$tables$mortality_glmm_history_emmeans_response
gt::gt(mme) %>%
  gt::tab_header(
    title = "Mortality: emmeans by prior stressor count",
    subtitle = "Tank-level binomial GLMM: cbind(dead, alive) ~ HistoryLevelNum + (1 | Tank.ID), type = response"
  ) %>%
  gt_fmt_numbers(mme, decimals = 3L, pvalue_threshold = 1e-4) %>%
  gt_bold_column_labels() %>%
  style_gt_significance(data_tbl = mme, sig_fill = "#C7EFCF")
```

***

## Table: mortality GLMM — polynomial trend (stress history)

```{r tbl-mort-glmm-poly, echo=FALSE}
mpoly <- mortinf$tables$mortality_glmm_history_poly_contrast
gt::gt(mpoly) %>%
  gt::tab_header(
    title = "Polynomial contrasts: mortality vs HistoryLevelNum",
    subtitle = "emmeans::contrast(emm, \"poly\") on HistoryLevelNum + (1 | Tank.ID) binomial GLMM"
  ) %>%
  gt_fmt_numbers(mpoly, decimals = 3L, pvalue_threshold = 1e-4) %>%
  gt_bold_column_labels() %>%
  style_gt_significance(data_tbl = mpoly, sig_fill = "#C7EFCF")
```

***

## Table: mortality pairwise contrasts (stress history levels, FDR)

```{r tbl-mort-glmm-history-pairs, echo=FALSE}
mpair <- mortinf$tables$mortality_glmm_history_pairs_fdr
gt::gt(mpair) %>%
  gt::tab_header(
    title = "Pairwise contrasts: mortality by prior stressor count",
    subtitle = "emmeans::contrast(..., \"pairwise\", adjust = \"fdr\") on HistoryLevelNum + (1 | Tank.ID) GLMM"
  ) %>%
  gt_fmt_numbers(mpair, decimals = 3L, pvalue_threshold = 1e-4) %>%
  gt_bold_column_labels() %>%
  style_gt_significance(data_tbl = mpair, sig_fill = "#C7EFCF")
```

***

## Table: infection prevalence by prior stressor history

```{r tbl-infection, echo=FALSE}
tbl_inf <- mortinf$tables$infection_prevalence_by_prior_stressor_history
if (!"n_survivors_sampled" %in% names(tbl_inf) && "n_fish" %in% names(tbl_inf)) {
  tbl_inf <- tbl_inf %>% dplyr::rename(n_survivors_sampled = n_fish)
}
tbl_inf <- tbl_inf %>%
  dplyr::select(
    dplyr::any_of(c(
      "History_Label",
      "n_survivors_sampled",
      "n_infected",
      "percent_infected",
      "mean_worm_burden",
      "total_worm_count",
      "mean_worm_burden_infected"
    ))
  )

tbl_inf %>%
  gt::gt() %>%
  gt::tab_header(
    title = "Infection in parasite-exposed fish (Day 60 survivors)",
    subtitle = "Unique sampled fish per prior stressor history; worm presence = Total.Worm.Count > 0"
  ) %>%
  gt::cols_label(
    History_Label = "Prior stressor history",
    n_survivors_sampled = "N survivors (sampled)",
    n_infected = "N infected",
    percent_infected = "Infected (%)",
    mean_worm_burden = "Mean worms (all fish)",
    total_worm_count = "Total worms",
    mean_worm_burden_infected = "Mean worms (infected only)"
  ) %>%
  gt_fmt_numbers(tbl_inf, decimals = 2L) %>%
  gt_bold_column_labels()
```

## Supplementary figures: mortality by stress history (levels)

### Mortality by prior stressor history (3 levels; history palette)

```{r fig-mort-history-levels, echo=FALSE, fig.cap="Mortality by prior stressor history (3 levels). Tank-level binomial GLMM: cbind(dead, alive) ~ HistoryLevel + (1 | Tank.ID). Colors use the stress-history palette (green/orange/purple)."}
include_png(mortinf$paths$figures$mortality_by_prior_stressor_history_levels_png)
```

### Mortality by prior stressor history × parasite exposure (history palette)

```{r fig-mort-history-levels-x-parasite, echo=FALSE, fig.cap="Mortality by prior stressor history × parasite exposure. Tank-level binomial GLMM: cbind(dead, alive) ~ HistoryLevel * Parasite + (1 | Tank.ID). Colors use the stress-history palette (green/orange/purple)."}
include_png(mortinf$paths$figures$mortality_by_prior_stressor_history_levels_x_parasite_png)
```

***



## Table: infection GLMM — marginal means (response scale)

```{r tbl-inf-emmeans, echo=FALSE}
em <- mortinf$tables$infection_glmm_history_emmeans_response
gt::gt(em) %>%
  gt::tab_header(
    title = "Infection prevalence: emmeans by prior stressor count",
    subtitle = "P+ Day 60 survivors; model: HistoryLevelNum + (1 | Tank.ID), type = response"
  ) %>%
  gt_bold_column_labels() %>%
  style_gt_significance(data_tbl = em, sig_fill = "#C7EFCF")
```

***

## Table: infection GLMM — polynomial trend (stress history)

```{r tbl-inf-poly, echo=FALSE}
pl <- mortinf$tables$infection_glmm_history_poly_contrast
gt::gt(pl) %>%
  gt::tab_header(
    title = "Polynomial contrasts: infection vs HistoryLevelNum",
    subtitle = "emmeans::contrast(emm, \"poly\") on HistoryLevelNum + (1 | Tank.ID) binomial GLMM"
  ) %>%
  gt_bold_column_labels() %>%
  style_gt_significance(data_tbl = pl, sig_fill = "#C7EFCF")
```

***

## Table: infection pairwise contrasts (stress history levels, FDR)

```{r tbl-inf-history-pairs, echo=FALSE}
hp <- mortinf$tables$infection_glmm_history_pairs_fdr
gt::gt(hp) %>%
  gt::tab_header(
    title = "Pairwise contrasts: infection by prior stressor count",
    subtitle = "emmeans::contrast(..., \"pairwise\", adjust = \"fdr\") on HistoryLevelNum + (1 | Tank.ID) GLMM"
  ) %>%
  gt_bold_column_labels() %>%
  style_gt_significance(data_tbl = hp, sig_fill = "#C7EFCF")
```

***
## Table: mortality GLMM joint tests (History × Parasite)

```{r tbl-mort-joint, echo=FALSE}
jt <- mortinf$tables$mortality_glmm_joint_tests
gt::gt(jt) %>%
  gt::tab_header(
    title = "Joint tests: mortality model",
    subtitle = "cbind(dead, alive) ~ HistoryLevelNum * Parasite + (1 | Tank.ID)"
  ) %>%
  gt_bold_column_labels() %>%
  style_gt_significance(data_tbl = jt, sig_fill = "#C7EFCF")
```

***

## Table: infection pairwise contrasts (P+ regimes, FDR)

```{r tbl-inf-pairs, echo=FALSE}
ip <- mortinf$tables$infection_glmm_treatment_pairs_fdr
gt::gt(ip) %>%
  gt::tab_header(
    title = "Pairwise contrasts: infection by Treatment (P+ Day 60)",
    subtitle = "emmeans pairs(..., adjust = 'fdr') on Treatment + (1 | Tank.ID) binomial GLMM"
  ) %>%
  gt_bold_column_labels() %>%
  style_gt_significance(data_tbl = ip, sig_fill = "#C7EFCF")
```

***

## Output paths

```{r paths, echo=FALSE}
mortinf$paths
```
