---
title: "Gene-taxa partial correlation network (display)"
subtitle: "06__Taxon-DEG-Mort — mirrors `Code/01__Analysis/06__Taxon-DEG-Mort.R`"
author: "Michael Sieler"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 6
params:
  rerun_analysis: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = isTRUE(params$rerun_analysis),
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 6,
  dpi = 300
)

start <- if (requireNamespace("knitr", quietly = TRUE) && !is.null(knitr::current_input())) {
  dirname(knitr::current_input(dir = TRUE))
} else {
  getwd()
}
init_rel <- file.path("Code", "00__Setup", "source_project_init.R")
root <- normalizePath(start, mustWork = TRUE)
while (!file.exists(file.path(root, init_rel))) {
  parent <- dirname(root)
  if (identical(parent, root)) {
    stop("Could not find ", init_rel, " above ", start)
  }
  root <- parent
}
source(file.path(root, init_rel))
sieler2026_source_initialize_project(root = root)

bundle_path <- here::here("Results", "06__Taxon-DEG-Mort", "Stats", "taxon_deg_mort__bundle.rds")

if (isTRUE(params$rerun_analysis)) {
  source(here::here("Code", "01__Analysis", "06__Taxon-DEG-Mort.R"))
}

if (!file.exists(bundle_path)) {
  stop(
    "Bundle not found: ", bundle_path, "\n",
    "Run `Code/01__Analysis/06__Taxon-DEG-Mort.R` from the project root, or knit with ",
    "`params$rerun_analysis: true`."
  )
}

taxon_deg <- readRDS(bundle_path)
per_taxon <- taxon_deg$tables$per_taxon_summary

# Resolve paths saved in bundles:
# - Prefer the stored path when it exists on this machine
# - Otherwise, treat it as relative to the repo root (here::here)
resolve_bundle_path <- function(p) {
  if (is.null(p) || !nzchar(p)) {
    return(NA_character_)
  }
  if (file.exists(p)) {
    return(p)
  }
  p2 <- tryCatch(here::here(p), error = function(e) NA_character_)
  if (!is.na(p2) && file.exists(p2)) {
    return(p2)
  }
  p3 <- file.path(root, p)
  if (file.exists(p3)) {
    return(p3)
  }
  p
}

include_png <- function(p) {
  p_res <- resolve_bundle_path(p)
  if (is.null(p_res) || !nzchar(p_res)) {
    return(invisible(NULL))
  }
  p_abs <- normalizePath(p_res, winslash = "/", mustWork = TRUE)
  knitr::include_graphics(p_abs, rel_path = FALSE)
}
```

This report loads `Results/06__Taxon-DEG-Mort/Stats/taxon_deg_mort__bundle.rds` produced by [`Code/01__Analysis/06__Taxon-DEG-Mort.R`](../01__Analysis/06__Taxon-DEG-Mort.R). **Spearman** correlations screen gene–genus pairs (both z-scored) within each exposure contrast vs **A- T- P-**; **nonparametric partial correlations** (`nptest::np.cor.test`) control for **Total.Worm.Count**. **FDR** (Benjamini–Hochberg) is applied to partial-correlation *p*-values (threshold **0.1**, matching the archived helper).

---

## Figure: bipartite partial-correlation network

```{r fig-bipartite, echo=FALSE, fig.cap="Significant partial correlations (FDR < 0.1), colored by partial r (red positive, blue negative). Genes (left) and genera (right); subset of top gene–taxon pairs for readability."}
include_png(taxon_deg$paths$figures$bipartite_png)
```

---

## Figure: partial correlation histogram (by contrast)

```{r fig-hist, echo=FALSE, fig.cap="Distribution of significant partial correlation estimates, faceted by base vs comparison regime; top genera by association count."}
include_png(taxon_deg$paths$figures$histogram_png)
```

---

## Figure: top 10 taxa (mean absolute partial correlation vs number of associations)

```{r fig-scatter, echo=FALSE, fig.cap="Top 10 genera ranked by count of FDR-significant partial-correlation edges (Benjamini-Hochberg FDR below 0.1 on nonparametric partial correlation p-values; infection burden adjusted via Total.Worm.Count). Spearman FDR below 0.05 screened gene-genus pairs before partial testing. x-axis: mean absolute partial correlation across significant edges. Point fill: mean signed partial correlation per genus (color bar -1 to 1). Blue = negative, red = positive. Focal genera (when present in the top 10): bold labels and ring; label colour matches sign of mean partial r."}
include_png(taxon_deg$paths$figures$top10_png)
```

---

## Figure: all significant taxa (full per-genus summary)

```{r fig-scatter-all, echo=FALSE, fig.cap="Every genus with at least one FDR-significant partial-correlation edge (same FDR 0.1 criterion as elsewhere in this module). Axes as in the top-10 figure. Smaller semi-transparent points for non-focal genera (no text labels). Focal genera (Shewanella, Culicoidibacter, Flavobacterium, Cetobacterium): larger points, hollow ring, bold ggrepel labels; label colour blue (negative mean partial r) or red (positive)."}
p_all <- taxon_deg$paths$figures$all_sig_png
if (!is.null(p_all) && !is.na(p_all) && nzchar(p_all)) {
  include_png(p_all)
} else {
  knitr::asis_output("_Figure path `all_sig_png` not in bundle (re-run `06__Taxon-DEG-Mort.R`)._")
}
```

---

## Figures: focal four genera (gene-set overlap for enrichment inputs)

Each panel summarizes **unique host gene symbols** with **BH FDR &lt; 0.1** on **nonparametric partial correlations** to genus abundance (infection burden adjusted via **Total.Worm.Count**), restricted to the four focal genera. **All**: every significant edge for that genus; **Positive** / **Negative**: genes with at least one significant edge of that sign. These are the same gene lists exported as `partial_correlation_edges__*.csv` and used as inputs to GO/KEGG enrichment in `Code/01__Analysis/07__FunctionalAnno.R`. Figures are written by `Code/01__Analysis/06__Taxon-DEG-Mort.R` (focal-four Venn helpers are inlined in that driver).

```{r fig-venn-focal-four-all, echo=FALSE, fig.cap="Gene overlap across focal four taxa (all FDR-significant partial-correlation edges per taxon)."}
p_va <- taxon_deg[["paths"]][["figures"]][["focal_four_venn_all"]]
if (is.null(p_va) || !nzchar(as.character(p_va))) {
  knitr::asis_output("_Venn path not in bundle (re-run `06__Taxon-DEG-Mort.R`)._")
} else {
  include_png(p_va)
}
```

```{r fig-venn-focal-four-positive, echo=FALSE, fig.cap="Gene overlap: genes with at least one positive partial correlation (FDR < 0.1) per taxon."}
p_vp <- taxon_deg[["paths"]][["figures"]][["focal_four_venn_positive"]]
if (is.null(p_vp) || !nzchar(as.character(p_vp))) {
  knitr::asis_output("_Venn path not in bundle._")
} else {
  include_png(p_vp)
}
```

```{r fig-venn-focal-four-negative, echo=FALSE, fig.cap="Gene overlap: genes with at least one negative partial correlation (FDR < 0.1) per taxon."}
p_vn <- taxon_deg[["paths"]][["figures"]][["focal_four_venn_negative"]]
if (is.null(p_vn) || !nzchar(as.character(p_vn))) {
  knitr::asis_output("_Venn path not in bundle._")
} else {
  include_png(p_vn)
}
```

### Table: overlap summary (Venn input gene sets)

```{r tbl-focal-four-venn-overlap-06, echo=FALSE}
tog <- taxon_deg[["table_focal_four_venn_gene_overlap"]]
if (is.null(tog)) {
  cat("Overlap summary not in bundle (zero significant edges, or re-run `06__Taxon-DEG-Mort.R`).\n")
} else {
  tog
}
```

---

## Table: per-genus summary (significant partial correlations)

```{r tbl-per-taxon, echo=FALSE}
if (nrow(per_taxon) > 0L) {
  if (!is.null(taxon_deg$table_per_taxon_summary)) {
    taxon_deg$table_per_taxon_summary
  } else {
    gt::gt(per_taxon)
  }
} else {
  cat("No significant partial correlations in bundle.")
}
```

---

## Table: bundle metadata

```{r tbl-meta, echo=FALSE}
tbl_meta <- tibble::tibble(
  Field = c("run_date", "script", "n_sig_partial_edges", "n_comparisons"),
  Value = c(
    taxon_deg$meta$run_date,
    taxon_deg$meta$script,
    as.character(taxon_deg$meta$n_sig_partial_edges),
    as.character(taxon_deg$meta$n_comparisons)
  )
)
gt::gt(tbl_meta) %>%
  gt::tab_header(title = "06__Taxon-DEG-Mort run summary")
```
